\textbf{ Mapping stellar surfaces\\ I: Degeneracies in the light curve problem }

### 2021-01-21 paper1/ms.tex
null space maths

% Title
\title{%
    \textbf{
        Mapping stellar surfaces\\
        I: Nullspace, degeneracies, and the unknowable
    }
}

@@

\keywords{time series analysis --- starspots}

@@

\begin{abstract}
    Abstract here.
\end{abstract}

@@

While the stellar surfaces are all distinct, containing between one (top)
and six (bottom) large dark spots,
\textbf{their rotational light curves are identical}
(lower panel). This is true even at infinite signal to
noise: the mapping from a stellar surface
to its rotational light curve is so degenerate that there exist an infinite number of solutions
to the inverse problem. This fact has been pointed out recently in different contexts
\citep[e.g.,][]{Cowan2013,Luger2019,Basri2020}, but it dates back at least to
\citet{Russell1906}, who demonstrated it by expanding the surface
intensity of a celestial body in terms of spherical harmonics
(see \S\ref{sec:main} and Figure~\ref{fig:ylms}). \citet{Russell1906} showed
that many of the modes comprising the intensity profile of a spherical
object are in the \emph{null space}, the set of surface features that have identically
zero effect on the light curve. In fact,
the \emph{vast majority} of the modes are in the null space for rotational
light curves of stars (\S\ref{sec:nullspace}). This is what allows us to construct pathological
scenarios like that shown in Figure~\ref{fig:degeneracies}, where the light curve
could be explained by any number of spots atop a heterogeneous bright background.

@@

Fortunately, although individual light curves are not very constraining,
we have \emph{a lot of them} thanks to missions like \emph{Kepler} and \emph{TESS}.
In this paper, we will thoroughly explore the power of ensemble analyses
in breaking the degeneracies inherent to the light curve mapping problem.
By jointly analyzing the light curves of many stars, we will show that it
is often possible to uniquely constrain statistical properties about their
surfaces. This idea was recently explored to some extent
in \citet{Morris2020}, who used ensemble analyses to derive constraints on
spot coverage areas as a function of stellar age (see \S\ref{sec:other-work}).

\subsection{The null space}
\label{sec:nullspace}

\begin{figure}[p!]
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/nullspace_0.pdf}
        \includegraphics[width=\linewidth]{figures/nullspace_1.pdf}
        \includegraphics[width=\linewidth]{figures/nullspace_2.pdf}
        \oscaption{nullspace}{%
            The nullspace decomposition.
            \label{fig:nullspace}
        }
    \end{centering}
\end{figure}

@@

>>>>

% Title
\title{%
    \textbf{
        Mapping stellar surfaces\\
        I: Degeneracies in the light curve problem
    }
}

@@

\keywords{time series analysis --- light curves --- stellar surfaces --- starspots}

@@

\begin{abstract}
    \xxx{Abstract here.}
\end{abstract}

@@

While the stellar surfaces are all distinct, containing between one (top)
and six (bottom) large dark spots,
\textbf{their rotational light curves are identical}
(lower panel). This is true even at infinite signal to
noise: the mapping from a stellar surface
to its rotational light curve is so degenerate that there exist an infinite number of solutions
to the inverse problem. This fact has been pointed out recently in different contexts
\citep[e.g.,][]{Cowan2013,Luger2019,Basri2020}, but it dates back at least to
\citet{Russell1906}, who demonstrated it by expanding the surface
intensity of a celestial body in terms of spherical harmonics.
(see Figure~\ref{fig:ylms}). \citet{Russell1906} showed
that many of the modes comprising the intensity profile of a spherical
object are in the \emph{null space}, the set of surface features that have identically
zero effect on the light curve. In fact,
the \emph{vast majority} of the modes are in the null space for rotational
light curves of stars (\S\ref{sec:nullspace}). This is what allows us to construct pathological
scenarios like that shown in Figure~\ref{fig:degeneracies}, where the light curve
could be explained by any number of spots atop a heterogeneous bright background.

\begin{figure}[t!]
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/ylms.pdf}
        \oscaption{ylms}{%
            The real spherical harmonics in the polar frame up to
            $l = 5$, where dark colors correspond to negative intensity
            and bright colors to positive intensity.
            Rows correspond to the degree $l$ and columns to
            the order $m$. The set of all spherical harmonics forms a
            complete, orthogonal basis on the sphere.
            \label{fig:ylms}
        }
    \end{centering}
\end{figure}

@@

% Fortunately, although individual light curves are not very constraining,
% we have \emph{a lot of them} thanks to missions like \emph{Kepler} and \emph{TESS}.
% In this paper, we will thoroughly explore the power of ensemble analyses
% in breaking the degeneracies inherent to the light curve mapping problem.
% By jointly analyzing the light curves of many stars, we will show that it
% is often possible to uniquely constrain statistical properties about their
% surfaces. This idea was recently explored to some extent
% in \citet{Morris2020}, who used ensemble analyses to derive constraints on
% spot coverage areas as a function of stellar age (see \S\ref{sec:other-work}).

\xxx{In this paper...}

\section{The null space}
\label{sec:nullspace}

\begin{figure}[p!]
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/nullspace_0.pdf}
        \\[1em]
        \includegraphics[width=\linewidth]{figures/nullspace_1.pdf}
        \\[1em]
        \includegraphics[width=\linewidth]{figures/nullspace_2.pdf}
        \oscaption{nullspace}{%
            The nullspace decomposition.
            \label{fig:nullspace}
        }
    \end{centering}
\end{figure}

@@

\xxx{A little linear algebra here:}

%
Linear mapping:
%
\begin{align}
    \mathbf{f} = \mathbf{A} \mathbf{y}
\end{align}
%
SVD:
%
\begin{align}
    \mathbf{A} = \mathbf{U} \mathbf{S} \mathbf{V}^\top
\end{align}
%
where $\mathbf{U}$ and $\mathbf{V}$ are
$(K \times N)$ orthogonal matrices
and $\mathbf{S}$ is a $(N \times N)$ diagonal matrix
containing the $R$ nonzero singular values as well as
$N - R$ zeros.
%
Write
%
\begin{align}
    \mathbf{A} =
    \begin{pmatrix}
        \mathbf{U}_\bullet & \mathbf{U}_\circ
    \end{pmatrix}
    \begin{pmatrix}
        \mathbf{S} &            \\
                   & \mathbf{0}
    \end{pmatrix}
    \begin{pmatrix}
        \mathbf{V}_\bullet^\top \\
        \mathbf{V}_\circ^\top
    \end{pmatrix}
\end{align}
%
Note that because $\mathbf{V}$ is orthogonal,
%
\begin{align}
    \mathbf{V}^\top \mathbf{V} = \mathbf{I}
\end{align}
%
and further
%
\begin{align}
    \mathbf{V}_\bullet^\top \mathbf{V}_\bullet = \mathbf{I}
\end{align}
%
and
\begin{align}
    \mathbf{V}_\circ^\top \mathbf{V}_\circ = \mathbf{I}
\end{align}
%
Then
%
\begin{align}
    \mathbf{f} & = \mathbf{A} \mathbf{y}
    \\
               & =
    \begin{pmatrix}
        \mathbf{U}_\bullet & \mathbf{U}_\circ
    \end{pmatrix}
    \begin{pmatrix}
        \mathbf{S} &            \\
                   & \mathbf{0}
    \end{pmatrix}
    \begin{pmatrix}
        \mathbf{V}_\bullet^\top \\
        \mathbf{V}_\circ^\top
    \end{pmatrix}
    \mathbf{y}
    \\
               & =
    \mathbf{U}_\bullet \mathbf{S} \mathbf{V}_\bullet^\top \mathbf{y}
    +
    \mathbf{U}_\circ \mathbf{0} \mathbf{V}_\circ^\top \mathbf{y}
    \\
               & =
    \mathbf{U}_\bullet \mathbf{S} (\mathbf{V}_\bullet^\top \mathbf{V}_\bullet) \mathbf{V}_\bullet^\top \mathbf{y}
    +
    \mathbf{U}_\circ \mathbf{0} (\mathbf{V}_\circ^\top \mathbf{V}_\circ) \mathbf{V}_\circ^\top \mathbf{y}
    \\
               & =
    (\mathbf{U}_\bullet \mathbf{S} \mathbf{V}_\bullet^\top) \mathbf{V}_\bullet \mathbf{V}_\bullet^\top \mathbf{y}
    +
    (\mathbf{U}_\circ \mathbf{0} \mathbf{V}_\circ^\top) \mathbf{V}_\circ \mathbf{V}_\circ^\top \mathbf{y}
    \\
               & =
    \mathbf{A} \mathbf{y}_\bullet
    +
    \mathbf{0} \mathbf{y}_\circ
\end{align}
%
where
%
\begin{align}
    \mathbf{y}_\bullet \equiv \mathbf{V}_\bullet \mathbf{V}_\bullet^\top \mathbf{y}
\end{align}
%
and
%
\begin{align}
    \mathbf{y}_\circ \equiv \mathbf{V}_\circ \mathbf{V}_\circ^\top \mathbf{y}
\end{align}
%
are the row space and null space components of $\mathbf{y}$.

\xxx{More on this.}
